### 2024-03-06 tex/examples.tex
[ correction ] finish chapter 8

The type of syntax descriptions depends on the type of usage annotations because
of variable binding.
For example, in the $\oc{\gr r}$-E rule of \cref{fig:lr-bunched}, the right
premise binds a new variable with annotation $\gr r$, where $\gr r$ is drawn
from the ambient posemiring.
The scaling connective also makes direct reference to the posemiring.
To produce a simplified syntax description, where usage constraints are
trivialised, we set the ambient posemiring to the 1-element $\mathbf0$
posemiring.
In contrast to syntax descriptions, even though types can contain usage
annotations, the type of types does not depend on the type of usage annotations.
This means that, in our simplified syntax, terms have types from the original
system even though variables have trivial usage annotations.
We define the $\mathbf0$ posemiring as follows, being careful to use the
0-field record type \AgdaRecord{$\top$} so that everything algebraic gets
solved by Agda's $\eta$-laws.
Indeed, in this very definition, all of the semiring operations and laws are
canonically inferred.

@@

@@

We can now use the elaborator to automatically infer the usage
annotations for the example at the end of \cref{sec:terms-of-system}.
This allows us to write:
\ExecuteMetaData[\PaperExamplestex]{cojoin}
We have instantiated the usage elaborator so that:
\AgdaField{0\#$^{-1}$} is a singleton on $\gr0$ and $\gr\omega$, and
empty on $\gr1$; \AgdaField{1\#$^{-1}$} is a singleton on $\gr1$ and
$\gr\omega$, and empty on $\gr0$; \AgdaField{+$^{-1}$} gives $\gr0
\mapsto [(\gr0,\gr0)]$, $\gr1 \mapsto [(\gr0,\gr1),(\gr1,\gr0)]$, and
$\gr\omega \mapsto [(\gr\omega,\gr\omega)]$; and \AgdaField{*$^{-1}$}
gives $(\gr\omega, \gr0) \mapsto [\gr0]$, $(\gr\omega, \gr1) \mapsto
[]$, and $(\gr\omega, \gr\omega) \mapsto [\gr\omega]$ (omitting
$(\gr0, \_)$ and $(\gr1, \_)$ cases for brevity). Note that we do not
consider splitting $\gr\omega$ up as, say, $\gr1 + \gr\omega$, because
this splitting would introduce more non-determinism but not allow any
more terms to be typed. As such, the only non-determinism comes when
we have variables annotated $\gr1$ and need to do an additive split,
like when we apply the \AgdaInductiveConstructor{!E} rule below. At
this point, the variable can become either $\gr0$-annotated in the
left subterm and $\gr1$-annotated on the right, or vice-versa. We will
find that, because the left subterm wants to use that variable, the
former choice will be rejected. The function \AgdaFunction{var\#} is a
convenience for converting statically known natural numbers,
representing De Bruijn \emph{levels}, into variable terms.

@@

The NbE model of types is given by \AgdaFunction{\_$\vDash$\_}.
It is defined by induction on the type.
The base type is interpreted by terms of base type, while
the function type is interpreted by the Kripke function space (essentially a
special case of \AgdaFunction{Kripke} with $\Delta = \gr rA$, though reusing
the \AgdaFunction{Kripke} definition causes difficulty for the termination
checker).
The interpretation of the base type additionally requires a \AgdaRecord{Lift}
of its universe level, because \AgdaFunction{$\Box^r$} produces a large type
(a type in \AgdaPrimitive{Set$_1$}).
The largeness of \AgdaFunction{$\Box^r$} comes from the largeness of
environments, and hence renamings, because of the relationally specified linear
map.

@@

@@

Standard denotational semantics falls out as a somewhat trivial case of
\AgdaRecord{Semantics}.
A lot of work is done in the generic traversal \AgdaFunction{semantics} to
maintain a $\V$-environment, where $\V$ is a reasonably variable-like semantic
family.
For a simple denotational semantics, we set $\V$ to be variables

@@

As a running example, I take the usage annotations to be the 4-element
variance posemiring (\cref{def:monotonicity-posemiring}).
I establish the property that all terms are monotonic in their free variables.
This monotonicity can be covariant or contravariant (or neither or both)
depending on the annotation of each free variable (respectively,
\AgdaInductiveConstructor{$\uparrow\uparrow$},
\AgdaInductiveConstructor{$\downarrow\downarrow$},
\AgdaInductiveConstructor{$\wn\wn$}, and
\AgdaInductiveConstructor{\textasciitilde\textasciitilde}).
This provides an additional example to those of \citet{AbelBernardy2020}.

>>>>

The type of syntax descriptions depends on the type of usage annotations because
of variable binding.
For example, in the $\oc{\gr r}$-E rule of \cref{fig:lr-bunched}, the right
premise binds a new variable with annotation $\gr r$, where $\gr r$ is drawn
from the ambient posemiring.
The scaling connective also makes direct reference to the posemiring.
To produce a simplified syntax description, where usage constraints are
trivialised, we set the ambient posemiring to the 1-element $\mathbf0$
posemiring.
In contrast to syntax descriptions, even though types can contain usage
annotations, the type of types does not depend on the type of usage annotations.
This means that, in the simplified syntax, terms have types from the original
system even though variables have trivial usage annotations.
I define the $\mathbf0$ posemiring as follows, being careful to use the
0-field record type \AgdaRecord{$\top$} so that everything algebraic gets
solved by Agda's $\eta$-laws.
Indeed, in this very definition, all of the semiring operations and laws are
canonically inferred.
This inference extends to entire contexts.
Recall that a context is a record containing a size and two functional vectors.
This means that, if we give the size and vector of types, the context is
determined up to definitional equality, because we know that the usage vector
is a function that always returns the unique usage annotation.

@@

\begin{example}\label{def:01w-inv}
  The example \AgdaRecord{NonDetInverses} I use in this section is for the
  linearity ($\instDILL$) semiring.
  The fields of the record are filled by the following functions.
  They implement all of the ``bottom-up'' laws, as defined in
  \cref{def:DILL-bottom-up}.

  To invert both \AgdaInductiveConstructor{0\#} and
  \AgdaInductiveConstructor{1\#}, we give one result whenever the respective
  relation (\AgdaBound{x} \AgdaDatatype{$\leq$} \AgdaInductiveConstructor{0\#}
  and \AgdaBound{x} \AgdaDatatype{$\leq$} \AgdaInductiveConstructor{1\#}) holds.
  This means giving a result both when \AgdaBound{x} is the annotation itself,
  and also when \AgdaBound{x} is \AgdaInductiveConstructor{$\omega$\#}.
  We are forced into this choice by the need to make $\gr\omega$-annotated
  variables both discardable and usable, without any intervening operations.

  \noindent
  \begin{minipage}[t]{0.5\textwidth}
    \ExecuteMetaData[\ZeroOneManytex]{0-inv}
  \end{minipage}
  \begin{minipage}[t]{0.5\textwidth}
    \ExecuteMetaData[\ZeroOneManytex]{1-inv}
  \end{minipage}

  In inverting addition, we return all possible results for
  \AgdaInductiveConstructor{0\#} and \AgdaInductiveConstructor{1\#}, but
  restrict the results for \AgdaInductiveConstructor{$\omega$\#} so as to avoid
  unnecessarily specialisation of $\gr\omega$-annotated variables into $0$- or
  $1$-annotated variables.
  Any variables annotated $\gr\omega$ will stay annotated $\gr\omega$ after
  applying rules using separating conjunction.

  \begin{samepage}
    \ExecuteMetaData[\ZeroOneManytex]{plus-inv}
  \end{samepage}

  Similarly to the case for addition, when inverting multiplication we apply
  the bottom-up restrictions.
  We enforce that multiplying by \AgdaInductiveConstructor{1\#} acts as a strict
  identity function, rather than allowing specialisation of
  \AgdaInductiveConstructor{$\omega$\#}.
  For \AgdaBound{x} \AgdaSymbol{=} \AgdaInductiveConstructor{0\#}, we know that
  \AgdaBound{z} has to be \AgdaInductiveConstructor{0\#} by annihilation.
  However, this much leaves \AgdaBound{y} completely unspecified, so we choose
  the most permissive result \AgdaInductiveConstructor{$\omega$\#}.
  For \AgdaBound{x} \AgdaSymbol{=} \AgdaInductiveConstructor{$\omega$\#},
  \AgdaInductiveConstructor{0\#}-annotated variables stay
  \AgdaInductiveConstructor{0\#}-annotated,
  \AgdaInductiveConstructor{$\omega$\#}-annotated variables stay
  \AgdaInductiveConstructor{$\omega$\#}-annotated, and
  \AgdaInductiveConstructor{1\#}-annotated variables are not allowed in the
  conclusion.
  The choices for
  \AgdaBound{x} \AgdaSymbol{=} \AgdaInductiveConstructor{$\omega$\#} correspond
  to how the $\oc$-introduction rule in DILL works.

  \begin{samepage}
    \ExecuteMetaData[\ZeroOneManytex]{mult-inv}
  \end{samepage}

  As argued by \cref{thm:DILL-bottom-up}, the bottom-up restrictions do not
  constrain the range of meaningfully different terms that can be produced.
\end{example}

@@

We can now use the elaborator to automatically infer the usage
annotations for the example at the end of \cref{sec:terms-of-system}.
Instantiating the elaborator to the \AgdaRecord{NonDetInverses} given in
\cref{def:01w-inv} allows us to write:

\ExecuteMetaData[\PaperExamplestex]{cojoin}

When elaborating this definition, the only non-determinism comes when
applying the \AgdaInductiveConstructor{!E}-rule with the
\AgdaInductiveConstructor{1\#}-annotated variable of type
\AgdaInductiveConstructor{!} \AgdaInductiveConstructor{$\omega$\#} %
\AgdaBound{A} in scope.
The non-determinism is resolved in the left subterm of this application, as the
application of the variable rule forces the annotation to become not
\AgdaInductiveConstructor{0\#}, hence \AgdaInductiveConstructor{1\#}, on the
left, and therefore become \AgdaInductiveConstructor{0\#} on the right.

Compare the one-line definition of \AgdaFunction{cojoin-!$\omega$} above with
the equivalent definition we saw in \cref{sec:terms-of-system} not using the
usage elaborator, in which we have to repeatedly specify both the usage contexts
and the (entirely obvious) proofs relating them.

\ExecuteMetaData[\HeavyItex]{cojoin-explicit}

%The function \AgdaFunction{var\#} is a
%convenience for converting statically known natural numbers,
%representing De Bruijn \emph{levels}, into variable terms.

@@

The NbE model of types is given by \AgdaFunction{\_$\vDash$\_}.
It is defined by induction on the type.
The base type is interpreted by terms of base type, while
the function type is interpreted by the Kripke function space (essentially a
special case of \AgdaFunction{Kripke} with $\Delta = \gr rA$)
\footnote{
  Using
  \AgdaFunction{Kripke} \AgdaFunction{\_$\vDash\_$} \AgdaFunction{\_$\vDash\_$}
  here causes difficulty for the termination checker because
  \AgdaFunction{Wrap}, and hence \AgdaFunction{Kripke}, does not reduce, leaving
  occurrences of \AgdaFunction{\_$\vDash\_$} not applied to smaller (or indeed
  any) arguments.
}.
The interpretation of the base type additionally requires a \AgdaRecord{Lift}
of its universe level, because \AgdaFunction{$\Box^r$} produces a large type
(a type in \AgdaPrimitive{Set$_1$}).
The largeness of \AgdaFunction{$\Box^r$} comes from the largeness of
environments, and hence renamings, because of the relationally specified linear
map.

@@

However, we can see by inspection that the algorithm given above is producing
normal forms and neutral forms in the places we would expect them to be.
At function type, \AgdaFunction{nbeReify} adds a $\lambda$-abstraction, around
the result of a recursive call, so if the base type values (terms) it receives
from the semantics are always normal, \AgdaFunction{nbeReify} will always
produce normal forms.
In fact, we expect the base type terms to be neutral forms.
We can see that they are neutral forms because we only ever pass
\AgdaFunction{nbeReflect} neutral forms, and the only larger terms it produces
are applications of the term it is given to normal forms coming from
\AgdaFunction{nbeReify}.
The traversal \AgdaFunction{evalSem} does not directly produce any terms; terms
(of base type) only appear via the environment.

@@

Standard denotational semantics falls out as a somewhat trivial case of
\AgdaRecord{Semantics}.
A lot of work is done in the generic traversal \AgdaFunction{semantics} to
maintain a $\V$-environment, where $\V$ is a reasonably variable-like semantic
family.
For a simple denotational semantics, we set $\V$ to be variables --- the minimal
family of values able to retain enough information to be meaningful in any
semantics --- and gradually build up an identity renaming ($\sqni$-environment)
as we go under binders.

@@

As a running example, I take the usage annotations to be the 4-element
variance posemiring (\cref{def:monotonicity-posemiring}).
I establish the property that all terms are monotonic in their free variables.
This monotonicity can be covariant or contravariant (or neither or both)
depending on the annotation of each free variable (respectively,
\AgdaInductiveConstructor{$\uparrow\uparrow$},
\AgdaInductiveConstructor{$\downarrow\downarrow$},
\AgdaInductiveConstructor{$\wn\wn$}, and
\AgdaInductiveConstructor{\textasciitilde\textasciitilde}).
This provides an additional example to those of \citet{AbelBernardy2020}.
The example includes the following type declaration, giving the carrier type of
the variance posemiring.
For the operations, refer to \cref{def:monotonicity-posemiring}.

\ExecuteMetaData[\Monotex]{WayUp}

### 2024-03-06 tex/syntax.tex
[ correction ] finish chapter 8

To see how the encodings of the operations work, let us look at the
\AgdaInductiveConstructor{`$\upmu\bullet$}-construct
($\tikz{\node (n) [circle,fill,minimum size=2mm,inner sep=0] {};%
  \draw[->] (n) to (2.12mm,-2.12mm); \draw[->] (n) to (-2.12mm,-2.12mm);%
  \draw[->] (0,3mm) to (n)}$).
It has two premises conjoined by $\sep$ (in fact, all premises in this language
are separated, with no use of sharing conjunction).
The first premise expects a simple value, corresponding to the input of the
operator.
The second premise is more complex, and represents the body/continuation of the
\texttt{let}-expression.
It binds two variables, corresponding to the two outputs, which can then be used
in later \texttt{let}-expressions or the outputs.
The bound variables all have type \AgdaInductiveConstructor{wire} and usage
annotation $\gr1$ (written \AgdaInductiveConstructor{u1} in Agda) to make them
behave linearly.

>>>>

To see how the encodings of the operations work, let us look at the
\AgdaInductiveConstructor{`$\upmu\bullet$}-construct
($\tikz{\node (n) [circle,fill,minimum size=2mm,inner sep=0] {};%
  \draw[->] (n) to (2.12mm,-2.12mm); \draw[->] (n) to (-2.12mm,-2.12mm);%
  \draw[->] (0,3mm) to (n)}$).
It has two premises conjoined by $\sep$ (in fact, all premises in this language
are separated, with no use of sharing conjunction).
The first premise expects a simple value, corresponding to the input of the
operator.
The second premise is more complex, and represents the body/continuation of the
\texttt{let}-expression.
It binds two variables, corresponding to the two outputs, which can then be used
in later \texttt{let}-expressions or the outputs.
The bound variables all have type \AgdaInductiveConstructor{wire} and usage
annotation $\gr1$ (written \AgdaInductiveConstructor{1\#} in Agda) to make them
behave linearly.
